---
title: "Aim4 Vitamin C Pathways and MoCA Scores"
authors: Cynthia Lin, Yanting Xiong
output: html_document
date: "2026-05-09"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Load Library

Initialize the environment with essential packages for microbiome data manipulation and tidy data processing.

```{r}
library(tidyverse)
library(ggplot2)
library(compositions)
library(vegan)
```

## Dataset Loading

Import Parkinson's metadata and MaAsLin2 results for vitamin C.

```{r Load datasets}
# Load Parkinson's metadata
parkinsons_metadata = read.csv("../../Datasets/Metadata/parkinsons_metadata_filt_cat.csv") %>% rename(SampleID = `X.SampleID`)

# Remove one PD patient with missing MoCA score
parkinsons_metadata_MoCA_NA_removed = parkinsons_metadata[parkinsons_metadata$SampleID != "F2501", ] 

# Load the Master Maaslin2 results for vitamin C
daa_results_VC = readRDS('../Aim_3 (Figure 3, Figure S3, Figure S4)/Vitamin_C/daa_master_results_with_L2FC.rds')

# Extract pathways that meet our significance thresholds (p-adjust < 0.05 & abs(log2fc) >= 1)
significant_df_VC = daa_results_VC %>%
  filter(p_adjust < 0.05, abs(log2fc) >= 1)

significant_pathways_VC = significant_df_VC$feature

```

## Load and Process Functional Pathway Data

Import raw pathway abundances, apply a CLR transformation, and subset for the significantly differentially abundant pathways identified in aim 3 for vitamin C.

```{r}
# Load raw pathway abundance table (rows = pathways, cols = samples)
pathway_counts = read.delim('../../QIIME2_outputs/path_abun_unstrat.tsv', row.names = 1)

# Manually remove PD patients with missing dietary antioxidant intake values and MoCA scores from the pathway abundance table
# The PD patient IDs are: F1452, F3436, F4446, F4484, F4660, F5362, F6222, F6413, F8334, F8922, F9523, F9556, F9856, F2501
pathway_counts[c("F1452", "F3436", "F4446", "F4484", "F4660", "F5362", "F6222", "F6413", "F8334", "F8922", "F9523", "F9556", "F9856", "F2501")] = NULL

# Calculate a pseudocount (half the minimum non-zero value across all the pathways)
min_nonzero = min(pathway_counts[pathway_counts > 0]) # 0.331428571428571

# Apply CLR transformation and transform the results into a data frame 
pathway_clr = clr(pathway_counts + min_nonzero/2) %>% as.data.frame()

# Subset the table for only significantly differentially abundant pathways identified in aim 3 and transpose so samples are rows
sig_pathways_vitamin_C = as.data.frame(t(pathway_clr[significant_pathways_VC, , drop = FALSE]))

# Calculate Spearman coefficient
spearman_VC = sig_pathways_vitamin_C %>% cor(method = "spearman") %>% 
  as.data.frame() %>% 
  rownames_to_column("PWY1") %>% 
  pivot_longer(cols = -PWY1, names_to = "PWY2", values_to = "cor") %>% 
  arrange(-cor)

# Select only those that are strongly correlated with a rho threshold of greater than 0.9
verycor_VC = spearman_VC %>% filter(cor > 0.9) %>% 
  filter(PWY1 != PWY2) # Remove self-correlations

# To avoid manually grouping them, since we have a lot of pathways that are highly correlated, we used a for loop approach:
# We will keep the pathway from each redundant pair that has the lowest p-value in Maaslin2.

# Create a lookup table for effect sizes from our Maaslin2 results
p_values_VC = significant_df_VC %>% select(feature, p_adjust) 

# Identify which pathways to drop
pathways_to_drop_VC = c()

for(i in 1:nrow(verycor_VC)) {
  p1 = verycor_VC$PWY1[i]
  p2 = verycor_VC$PWY2[i]
  
  # Only process if we haven't already dropped one of them in a previous loop
  if(!(p1 %in% pathways_to_drop_VC) && !(p2 %in% pathways_to_drop_VC)){
    
    # Compare their p-values
    p_value_p1 = p_values_VC$p_adjust[p_values_VC$feature == p1]
    p_value_p2 = p_values_VC$p_adjust[p_values_VC$feature == p2]
    
    # Drop the one with the higher p-value
    if(p_value_p1 > p_value_p2) {
      pathways_to_drop_VC = c(pathways_to_drop_VC, as.character(p1))
    } else {
      pathways_to_drop_VC = c(pathways_to_drop_VC, as.character(p2))
    }
  }
}

pathways_to_drop_VC = unique(pathways_to_drop_VC)

# Create our final, non-redundant list
final_pathways_VC = setdiff(significant_pathways_VC, pathways_to_drop_VC)
# Started with 37 pathways, kept 17 non-redundant pathways.

# Update our working matrix to only include the non-redundant pathways
sig_pathways_vitamin_C_filtered = sig_pathways_vitamin_C %>% select(all_of(final_pathways_VC))

```

## PCoA analysis

Use Aitchison distance to run PCoA analysis and visualize using ggplot2.

```{r}
# Compute dissimilarity matrix - Aitchison distance on CLR data
aitchison_dist_VC = dist(sig_pathways_vitamin_C_filtered, method = "euclidean")
# View(as.matrix(euclidean_dist_VC)) # Each pairwise comparison included

# Run PCoA analysis
pcoa_VC = cmdscale(aitchison_dist_VC, k = 2, eig = TRUE)

# Extract coordinates and variance explained
points_VC = as.data.frame(pcoa_VC$points)

colnames(points_VC) = c("PCoA1", "PCoA2")

points_VC$SampleID = rownames(points_VC)

points_VC = merge(
  points_VC,
  parkinsons_metadata_MoCA_NA_removed[, c("SampleID", "moca_total", "Sex", "Age", "entacapone")],
  by = "SampleID"
)

variance_explained_VC = round(100 * pcoa_VC$eig / sum(pcoa_VC$eig), 2)

# Visualization with ggplot2
pcoa_moca_plot_VC = ggplot(points_VC, aes(x = PCoA1, y = PCoA2, color = moca_total)) +
  geom_point(size = 3) +
  stat_ellipse(level = 0.95, linetype = 2, alpha = 0.3) +
  labs(
    x = paste0("PCoA1 (", variance_explained_VC[1], "%)"),
    y = paste0("PCoA2 (", variance_explained_VC[2], "%)"),
    color = "MoCA Score"
 ) +
  theme_minimal() +
  theme_classic(base_size = 14) +
  theme(axis.title = element_text(face = "bold")) 
pcoa_moca_plot_VC
  
# Export the ordination plot
ggsave("../../Results/Figures/Aim_4 (Figure_2, Figure_S5)/Pathways/VC_PCoA_NonRedundant_Pathway_vs_MoCA.png", plot = pcoa_moca_plot_VC, width = 10, height = 7, dpi = 800)

```

## PERMANOVA test

PERMANOVA test is performed to determine the statistical significance of the correlation.

```{r}
# Match the Parkinson's metadata to the distance matrix sample order --> reorders metadata so rows match the distance matrix order exactly
parkinsons_metadata_match_VC = parkinsons_metadata_MoCA_NA_removed[match(labels(aitchison_dist_VC), # Gets the sample names from the distance matrix
                                                                         parkinsons_metadata_MoCA_NA_removed$SampleID), ]

# Remove rows with missing values (10 PD patients with missing entacapone intake information are removed)
parkinsons_metadata_match_VC = parkinsons_metadata_match_VC[complete.cases(parkinsons_metadata_match_VC[, c("moca_total", "Age", "Sex", "entacapone")]), ]

# Save sample IDs of kept samples
keep_VC = parkinsons_metadata_match_VC$SampleID

# Subset the distance matrix
dist_mat_VC = as.dist(as.matrix(aitchison_dist_VC)[ # Converts distance object into a full square matrix
  match(keep_VC, labels(aitchison_dist_VC)), 
  match(keep_VC, labels(aitchison_dist_VC)) 
])

# Run PERMANOVA test to assess whether variation in the pathway distance matrix is associated with MoCA score after accounting for age, sex, and entacapone use
set.seed(421)

permanova_test = adonis2(
  dist_mat_VC ~ moca_total + Age + Sex + entacapone,
  data = parkinsons_metadata_match_VC,
  by = "margin",
  permutations = 999
)
permanova_test # p-value of moca_total is 0.448

```
