---
title: "aim4_pathway_vitamin_C_draft"
output: html_document
date: "2026-05-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Load Library
Initialize the environment with essential packages for microbiome data manipulation and tidy data processing.

```{r}
library(tidyverse)
library(phyloseq)
library(ggplot2)
```

## Dataset Loading
Import raw feature tables, taxonomy assignments, and clinical metadata.

```{r Load datasets}
# Load clinical and DAI metadata
metadata = read.csv("../../Datasets/Metadata/parkinsons_metadata_filt_cat.csv", row.names = 1)
metadata$SampleID = rownames(metadata)

# Load the Master Maaslin2 results (must include p_adjust and manual log2fc)
daa_results = readRDS('../Aim_3 (Figure 3, Figure S3, Figure S4)/Vitamin_C/daa_master_results_with_L2FC.rds')

# Dynamically extract pathway IDs that met our significance thresholds
significant_df = daa_results %>%
  filter(p_adjust < 0.05, abs(log2fc) >= 1)

significant_pathways = significant_df$feature

```

## Load and Process Functional Pathway Data
Import raw pathway abundances, convert to relative abundance, subset for our significant pathways, and apply a log10 transformation.

```{r}
# Load raw pathway abundance table (rows = pathways, cols = samples)
pathway_counts = read.delim('../../QIIME2_outputs/path_abun_unstrat.tsv', row.names = 1)

# Convert pathway abundance to relative abundance to normalize the counts
pathway_relab = sweep(
  pathway_counts, 
  2, 
  colSums(pathway_counts),
  `/`
  )

# Subset the table for only our significant pathways and transpose so samples are rows
sub_pathways = as.data.frame(t(pathway_relab[significant_pathways, , drop = FALSE]))

# Calculate a dynamic pseudocount (half the minimum non-zero value across these specific pathways)
min_nonzero = min(sub_pathways[sub_pathways > 0]) 

# Apply CLR transformation
pathway_clr = compositions::clr(sub_pathways + min_nonzero/2) %>% as.data.frame()

# Calculate Spearman coefficient
spearman = pathway_clr %>% cor(method = "spearman") %>% 
  as.data.frame() %>% 
  rownames_to_column('PWY1') %>% 
  pivot_longer(cols = -PWY1, names_to = 'PWY2', values_to = 'cor') %>% 
  arrange(-cor)

# Select only those that are strongly correlated
verycor = spearman %>% filter(cor > 0.9) %>% 
  filter(PWY1!=PWY2)

```