---
title: "Aim4_beta-carotene_pathway_draft"
authors: Cynthia Lin, Yanting Xiong
output: html_document
date: "2026-05-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Library
Initialize the environment with essential packages for microbiome data manipulation and tidy data processing.

```{r}
library(tidyverse)
library(phyloseq)
library(ggplot2)
library(compositions)
```

## Dataset Loading
Import raw feature tables, taxonomy assignments, and clinical metadata.

```{r Load datasets}
# Load clinical and DAI metadata
parkinsons_metadata = read.csv("../../Datasets/Metadata/parkinsons_metadata_filt_cat.csv") %>% rename(SampleID = `X.SampleID`)

# Because we will be controlling for entacapone in downstream analyses and it contains NA values, they need to be removed from the filtered metadata
parkinsons_metadata_entacapone_NA_removed = parkinsons_metadata %>% filter(!is.na(entacapone))
# 10 PD individuals are removed after this filtering step because they do not have entacapone intakes recorded

# Load the Master Maaslin2 results for vitamin C
daa_results_BC = readRDS('../Aim_3 (Figure 3, Figure S3, Figure S4)/Beta-caro/daa_master_results_with_L2FC.rds')

# Extract pathways that meet our significance thresholds (p-adjust < 0.05 & abs(log2fc) >= 1)
significant_df_BC = daa_results_BC %>%
  filter(p_adjust < 0.05, abs(log2fc) >= 1)

significant_pathways_BC = significant_df_BC$feature

```

## Load and Process Functional Pathway Data
Import raw pathway abundances, apply a CLR transformation, and subset for the significantly differentially abundant pathways identified in aim 3 for beta-carotene.

```{r}
# Load raw pathway abundance table (rows = pathways, cols = samples)
pathway_counts = read.delim('../../QIIME2_outputs/path_abun_unstrat.tsv', row.names = 1)

# Calculate a pseudocount (half the minimum non-zero value across all the pathways)
min_nonzero = min(pathway_counts[pathway_counts > 0]) # 0.331428571428571

# Apply CLR transformation and transform the results into a data frame 
pathway_clr = clr(pathway_counts + min_nonzero/2) %>% as.data.frame()

# Subset the table for only significantly differentially abundant pathways identified in aim 3 and transpose so samples are rows
sig_pathways_beta_carotene = as.data.frame(t(pathway_clr[significant_pathways_BC, , drop = FALSE]))

# Calculate Spearman coefficient
spearman_BC = sig_pathways_beta_carotene %>% cor(method = "spearman") %>% 
  as.data.frame() %>% 
  rownames_to_column("PWY1") %>% 
  pivot_longer(cols = -PWY1, names_to = "PWY2", values_to = "cor") %>% 
  arrange(-cor)

# Select only those that are strongly correlated with a rho threshold of greater than 0.9
verycor_BC = spearman_BC %>% filter(cor > 0.9) %>% 
  filter(PWY1 != PWY2) # Remove self-correlations

# To avoid manually grouping them, since we have a lot of pathways that are highly correlated, we used a for loop approach:
# We will keep the pathway from each redundant pair that has the lowest p-value in Maaslin2.

# Create a lookup table for effect sizes from our Maaslin2 results
p_values_BC = significant_df_BC %>% select(feature, p_adjust) 

# Identify which pathways to drop
pathways_to_drop_BC = c()

for(i in 1:nrow(verycor_BC)) {
  p1 = verycor_BC$PWY1[i]
  p2 = verycor_BC$PWY2[i]
  
  # Only process if we haven't already dropped one of them in a previous loop
  if(!(p1 %in% pathways_to_drop_BC) && !(p2 %in% pathways_to_drop_BC)){
    
    # Compare their p-values
    p_value_p1 = p_values_BC$p_adjust[p_values_BC$feature == p1]
    p_value_p2 = p_values_BC$p_adjust[p_values_BC$feature == p2]
    
    # Drop the one with the higher p-value
    if(p_value_p1 > p_value_p2) {
      pathways_to_drop_BC = c(pathways_to_drop_BC, as.character(p1))
    } else {
      pathways_to_drop_BC = c(pathways_to_drop_BC, as.character(p2))
    }
  }
}

pathways_to_drop_BC = unique(pathways_to_drop_BC)

# Create our final, non-redundant list
final_pathways_BC = setdiff(significant_pathways_BC, pathways_to_drop_BC) 
# Started with 21 pathways, kept 10 non-redundant pathways.

# Update our working matrix to only include the non-redundant pathways
sig_pathways_beta_carotene_filtered = sig_pathways_beta_carotene %>% select(all_of(final_pathways_BC))

```

## PCoA analysis
Use Euclidean distance to run PCoA analysis.

```{r}
# Euclidean distance on CLR data
euclidean_dist_BC = dist(sig_pathways_beta_carotene_filtered, method = "euclidean")
# View(as.matrix(euclidean_dist_BC)) # Each pairwise comparison included

# Run PCoA analysis
pcoa_BC = cmdscale(euclidean_dist_BC, k = 2, eig = TRUE)

# Calculate variance explained by Axis 1
variance_explained_BC = pcoa_BC$eig / sum(pcoa_BC$eig) * 100
variance_pc1_BC = variance_explained_BC[1]

# Extract Axis 1 (our new combined metric) and format for merging
pcoa_df_BC = data.frame(SampleID = rownames(pcoa_BC$points),
                        PCoA_Axis1_BC = pcoa_BC$points[, 1])

# Merge with clinical metadata
parkinsons_meta_merged = merge(parkinsons_metadata_entacapone_NA_removed, pcoa_df_BC, by = "SampleID")

```

## Linear Regression Model
Test the association between the combined pathway metric (PC1) and MoCA scores while controlling for confounders.

```{r}
# Predict MoCA based on PCoA_Axis1, adjusting for Age, Sex, and entacapone
model = lm(moca_total ~ PCoA_Axis1_BC + Age + Sex + entacapone, data = parkinsons_meta_merged)

summary(model)
```

## Visualization of Linear Regression Model
Generate a scatter plot with a linear regression line for the final manuscript.

```{r}
pcoa_moca_plot = ggplot(parkinsons_meta_merged, aes(x = PCoA_Axis1_BC, y = moca_total)) +
  geom_point(size = 3, alpha = 0.7) + 
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = paste0("Combined Pathway Score (PCoA Axis 1)\nExplains ", round(variance_pc1_BC, 1), "% of Variance"),
       y = "MoCA Score") +
  theme_classic(base_size = 16) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

pcoa_moca_plot

ggsave("../../Results/Figures/Aim_4 (Figure_2, Figure_S5)/Pathways/BC_PCoA_NonRedundant_Pathway_vs_MoCA.png", plot = pcoa_moca_plot, width = 10, height = 7, dpi = 300)

```
