---
title: "pathway_testing"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Library
Initialize the environment with essential packages for microbiome data manipulation and tidy data processing.

```{r}
library(tidyverse)
library(phyloseq)
library(ggplot2)
```
## Dataset Loading
Import raw feature tables, taxonomy assignments, and clinical metadata.
```{r Load datasets}
# Load clinical and DAI metadata
metadata = read.csv("../../Datasets/Metadata/parkinsons_metadata_filt_cat.csv", row.names = 1)
metadata$SampleID = rownames(metadata)

# Load the Master Maaslin2 results (must include p_adjust and manual log2fc)
daa_results = readRDS('../Aim_3 (Figure 3, Figure S3, Figure S4)/Vitamin_C/daa_master_results_with_L2FC.rds')

# Dynamically extract pathway IDs that met our significance thresholds
significant_df = daa_results %>%
  filter(p_adjust < 0.05, abs(log2fc) >= 1)

significant_pathways = significant_df$feature

```

## Load and Process Functional Pathway Data
Import raw pathway abundances, convert to relative abundance, subset for our significant pathways, and apply a log10 transformation.

```{r}
# Load raw pathway abundance table (rows = pathways, cols = samples)
pathway_counts = read.delim('../../QIIME2_outputs/path_abun_unstrat.tsv', row.names = 1)

# Convert to relative abundance (column-wise division by column sums)
pathway_relab = sweep(pathway_counts, 2, colSums(pathway_counts), `/`)

# Subset the table for ONLY our significant pathways and transpose so samples are rows
sub_pathways = as.data.frame(t(pathway_relab[significant_pathways, , drop = FALSE]))

# Calculate a dynamic pseudocount (half the minimum non-zero value across these specific pathways)
min_nonzero = min(sub_pathways[sub_pathways > 0])

# Apply Log10 transformation to all selected pathway columns simultaneously
sub_log = sub_pathways %>%
  mutate(across(everything(), ~ log10(.x + (min_nonzero / 2))))

sub_log$SampleID = rownames(sub_log)

```

## Metadata Integration
Join the clinical metadata with the log-transformed pathway abundances

```{r}
# Merge by SampleID
meta_merged = merge(metadata, sub_log, by = "SampleID")

# Verify the merge
head(meta_merged)
```

## Principal Component Analysis (PCA)
Collapse the multiple significant pathways into a single "Combined Metric" (PC1) to avoid multicollinearity in our linear model.

```{r}
# Isolate just the numeric pathway columns for the PCA matrix
pathway_matrix = meta_merged %>% 
  select(all_of(significant_pathways))

# Run PCA - CRITICAL: scale. = TRUE ensures high-abundance pathways don't dominate the score
pca_result = prcomp(pathway_matrix, center = TRUE, scale. = TRUE)

# Calculate variance explained by PC1
pca_summary = summary(pca_result)
variance_explained_pc1 = pca_summary$importance[2, 1] * 100

print(paste("PC1 explains", round(variance_explained_pc1, 2), "% of the variance among the selected pathways."))

# Append the PC1 scores back to our metadata table for the linear model
meta_merged$Combined_Pathway_PC1 = pca_result$x[, 1]
```

## Check Pathway Loadings
Identify which specific pathways are driving the PC1 score.

```{r}
# Extract the rotation (loadings) matrix for PC1
pc1_loadings = as.data.frame(pca_result$rotation[, 1, drop = FALSE])
colnames(pc1_loadings) = c("Loading")
pc1_loadings$Pathway = rownames(pc1_loadings)

# Sort to see the strongest drivers
pc1_loadings = pc1_loadings %>% arrange(desc(abs(Loading)))
print("Top pathways driving the Combined Metric (PC1):")
head(pc1_loadings)
```
## Linear Regression Model
Test the association between the combined pathway metric (PC1) and MoCA scores while controlling for confounders.

```{r}
# Predict MoCA based on PC1, adjusting for Age, Sex, and entacapone
model = lm(moca_total ~ Combined_Pathway_PC1 + Age + Sex + entacapone, data = meta_merged)

summary(model)
```
## Visualization of Linear Regression Model
Generate a scatter plot with a linear regression line for the final manuscript.

```{r}
# Plot PC1 against MoCA Total
pca_moca_plot = ggplot(meta_merged, aes(x = Combined_Pathway_PC1, y = moca_total)) +
  geom_point(size = 3, alpha = 0.7) + 
  geom_smooth(method = "lm", color = "black", se = TRUE) +
  scale_color_viridis_c(name = "Age") +
  labs(x = paste0("Combined Pathway Score (PC1)\nExplains ", round(variance_explained_pc1, 1), "% of Variance"),
       y = "MoCA Score") +
  theme_classic(base_size = 18) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

pca_moca_plot

# Export the plot

ggsave("../../Results/Figures/Aim_4 (Figure_2, Figure_S5)/Pathways/PCA_Combined_Pathway_vs_MoCA.png", 
       plot = pca_moca_plot, width = 10, height = 7, dpi = 600)
```

