---
title: "VC_pathway_test1"
output: html_document
author: Cynthia Lin
date: "2026-05-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction
This script investigates whether a coordinated shift in Vitamin C-associated metabolic pathways predicts cognitive performance (MoCA scores). Based on standard microbiome methodology, we first remove highly redundant pathways (Spearman rho > 0.9) to prevent artificial skewing. We then use Principal Coordinates Analysis (PCoA) on the non-redundant pathways to create a single composite metric (Axis 1), which is tested against MoCA scores while adjusting for clinical covariates.

## Load Library
Initialize the environment with essential packages for microbiome data manipulation and tidy data processing.

```{r}
library(tidyverse)
library(phyloseq)
library(ggplot2)
library(vegan)
```
## Dataset Loading
Import raw feature tables, taxonomy assignments, and clinical metadata.
```{r Load datasets}
# Load clinical and DAI metadata
metadata = read.csv("../../Datasets/Metadata/parkinsons_metadata_filt_cat.csv", row.names = 1)
metadata$SampleID = rownames(metadata)

# Load the Master Maaslin2 results (must include p_adjust and manual log2fc)
daa_results = readRDS('../Aim_3 (Figure 3, Figure S3, Figure S4)/Vitamin_C/daa_master_results_with_L2FC.rds')

# Dynamically extract pathway IDs that met our significance thresholds
significant_df = daa_results %>%
  filter(p_adjust < 0.05, abs(log2fc) >= 1)

significant_pathways = significant_df$feature

```

## Load Functional Pathway Data
Import raw pathway abundances, subset for our significant pathways

```{r}
# Load metadata
metadata = read.csv("../../Datasets/Metadata/parkinsons_metadata_filt_cat.csv", row.names = 1)
metadata$SampleID = rownames(metadata)

# Load Maaslin2 Master Results
daa_results = readRDS('../Aim_3 (Figure 3, Figure S3, Figure S4)/Vitamin_C/daa_master_results_with_L2FC.rds')

# Extract ONLY the significant pathway IDs
significant_df = daa_results %>% filter(p_adjust < 0.05, abs(log2fc) >= 1)
significant_pathways = significant_df$feature

# Load raw pathway abundance table
pathway_counts = read.delim('../../QIIME2_outputs/path_abun_unstrat.tsv', row.names = 1)

# Subset pathway table for only our significant pathways and transpose
pathway_matrix = as.data.frame(t(pathway_counts[significant_pathways, , drop = FALSE]))

```

## Redundancy Filtering (Spearman Correlation > 0.9)
Many MetaCyc pathways share the same underlying gene reads. We use Spearman correlation to identify and group highly redundant pathways.

```{r}
# Calculate Spearman correlation matrix between pathways
cor_matrix = cor(pathway_matrix, method = "spearman")

# Convert matrix to a long dataframe for easier filtering
cor_df = as.data.frame(as.table(cor_matrix)) %>%
  rename(PWY1 = Var1, PWY2 = Var2, rho = Freq) %>%
  filter(PWY1 != PWY2) %>%  # Remove self-correlations
  arrange(desc(rho))

# Filter for highly correlated pathway pairs (rho > 0.9)
redundant_pairs = cor_df %>% filter(rho > 0.9)

# To avoid manually grouping them, since we have a lot of pathways that are highly correlated, we used a for loop approach:
# We will keep the pathway from each redundant pair that had the highest absolute Log2FC in Maaslin2.

# Create a lookup table for effect sizes from our Maaslin2 results
effect_sizes = significant_df %>% select(feature, log2fc) %>% mutate(abs_l2fc = abs(log2fc))

# Identify which pathways to DROP
pathways_to_drop = c()

for(i in 1:nrow(redundant_pairs)) {
  p1 = redundant_pairs$PWY1[i]
  p2 = redundant_pairs$PWY2[i]
  
  # Only process if we haven't already dropped one of them in a previous loop
  if(!(p1 %in% pathways_to_drop) && !(p2 %in% pathways_to_drop)){
    
    # Compare their effect sizes
    eff_p1 = effect_sizes$abs_l2fc[effect_sizes$feature == p1]
    eff_p2 = effect_sizes$abs_l2fc[effect_sizes$feature == p2]
    
    # Drop the one with the smaller effect size
    if(eff_p1 > eff_p2) {
      pathways_to_drop = c(pathways_to_drop, as.character(p2))
    } else {
      pathways_to_drop = c(pathways_to_drop, as.character(p1))
    }
  }
}

pathways_to_drop = unique(pathways_to_drop)

# Create our final, non-redundant list
final_pathways = setdiff(significant_pathways, pathways_to_drop)
# Started with 37 pathways. Kept 18 non-redundant pathways.

# Update our working matrix to only include the non-redundant pathways
pathway_matrix_filtered = pathway_matrix %>% select(all_of(final_pathways))
```

## Data Transformation and PCoA
Following microbiome standards for compositional data, we convert to relative abundance, apply a small pseudocount, and use Bray-Curtis distance for the PCoA.
```{r}
# Identify if any samples have a total sum of 0 for our selected pathways
zero_sums = rowSums(pathway_matrix_filtered) == 0

if(any(zero_sums)) {
  pathway_matrix_filtered = pathway_matrix_filtered[!zero_sums, ]
}

# Convert to relative abundance
pathway_relab = sweep(pathway_matrix_filtered, 1, rowSums(pathway_matrix_filtered), `/`)

# Calculate Bray-Curtis distance matrix
# (vegan's vegdist requires numeric matrices)
bc_dist = vegdist(pathway_relab, method = "bray")

# Run Principal Coordinates Analysis (PCoA / MDS)
pcoa_result = cmdscale(bc_dist, k = 2, eig = TRUE)

# Calculate variance explained by Axis 1
variance_explained = pcoa_result$eig / sum(pcoa_result$eig) * 100
variance_pc1 = variance_explained[1]

# Extract Axis 1 (our new combined metric) and format for merging
pcoa_df = data.frame(SampleID = rownames(pcoa_result$points),
                     PCoA_Axis1 = pcoa_result$points[, 1])

# Merge with clinical metadata
meta_merged = merge(metadata, pcoa_df, by = "SampleID")
```

## Linear Regression Model
Test the association between the combined pathway metric (PC1) and MoCA scores while controlling for confounders.
```{r}
# Predict MoCA based on PCoA_Axis1, adjusting for Age, Sex, and entacapone
model = lm(moca_total ~ PCoA_Axis1 + Age + Sex + entacapone, data = meta_merged)

summary(model)
```

## Visualization of Linear Regression Model
Generate a scatter plot with a linear regression line for the final manuscript.

```{r}
pcoa_moca_plot = ggplot(meta_merged, aes(x = PCoA_Axis1, y = moca_total)) +
  geom_point(size = 3, alpha = 0.7) + 
  geom_smooth(method = "lm", color = "black", se = TRUE) +
  labs(x = paste0("Combined Pathway Score (PCoA Axis 1)\nExplains ", round(variance_pc1, 1), "% of Variance"),
       y = "MoCA Score",
       title = "Association of Non-Redundant Pathway Shift with MoCA") +
  theme_classic(base_size = 16) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

pcoa_moca_plot

ggsave("../../Results/Figures/Aim_4 (Figure_2, Figure_S5)/Pathways/PCoA_NonRedundant_Pathway_vs_MoCA.png", plot = pcoa_moca_plot, width = 10, height = 7, dpi = 300)
```

